Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2019-03-19 doc/network.tex
Worked through time; space next

\jd{Consider rewriting to use early-late as the first-level structure. Early is the part that people focus on most. We could even throw in censorship.}
\swp{Not sure what you mean... I tried to focus on early-late throughout the manuscript}
Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
As epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution.
When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{champredon2015intrinsic}.

In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, these measurements are \emph{censored} (or right-censored) at the observation time.
These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.

Observed generation intervals also vary across space.
In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic. 
This means that contacts made late in infection are more likely to be ``masked'' due to infecious contacts that were made earlier.
Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

In this study, we explore the spatiotemporal variation in the observed generation-interval distributions obtained from contact tracing.
We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
\swp{Come back and fix this}

@@

Generation-interval distributions are often viewed from a population-level pespective but we can distinguish population-level distributions from individual-level distributions (\fref{indpop}); 
making this distinction clear will be particularly useful when we discuss spatial components.
An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$ \citep{svensson2007note, svensson2015influence}.
The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
\jd{Do we need to worry about assumptions for the rest of this \P? Once we have an aspect space we need to assume that individual properties are independent of risk of infection. Also, since that's wrong on our networks, do we have to figure out why we're approximately saved? I guess because of scaling?}
\swp{What do you mean by individual properties or even scaling?}

There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
The basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
\begin{equation}
\RR_0 = \int K(t).
\end{equation}
The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
\begin{equation}
g(\tau) = \frac{K(t)}{\RR_0}.
\end{equation}

We assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
\begin{equation}
i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds.
\end{equation}
This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
Over a period of time where $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
\begin{equation}
\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
\end{equation}

@@

\jd{Mathematically, early censored and early backward intervals are the same, but conceptually they're just different. In particular, censoring occurs even with forward tracing.}
\jd{During exponential growth, the theoretical distributions are the same; maybe we should do a tiny bit of that math here so that we can talk more clearly about it.}
\jd{We need to think about exposition. Maybe simple formulae for all of our interval types, both in general and under the assumption of exponential growth.}
\jd{We should be saying that backward $\to$ censored in the limit at the same time we talk about the association below.}
\swp{I tried}

Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
Several studies have associated the backward generation-interval distribution with the observed generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}, but the backward generation-interval distribution takes into account infections that are only a single generation apart from the focal time.
Instead, contact tracing can be performed over multiple generations, tracing all the way back to the primary case.
The backward generation-interval distribution is mathematically equivalent to the censored generation-interval distribution only during the exponential growth phase, when depletion of the susceptible population is negligible.
However, the exponential growth assumption breaks down rapidly, and it is important to be able to conceptually distinguish the censored generation-interval from the backward generation-interval.
\jd{I hope that none of these papers actually make the mistake that you imply, it's all about the limit.}
\swp{Sure, in the limit they're all same. But we're often not in the limit and contact tracing can be performed beyond the limit. We can make this point.}

The number of infection occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
\begin{equation}
i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-x) g(x) d\tau}
\end{equation}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by
\begin{equation}\label{eq:obsg}
c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(\tau) \int_0^t i(s-\tau) S(s) ds d\tau}
\end{equation}
Alternatively, the censored generatio-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) S(s) ds.
\end{equation}
Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$.

>>>>

Early in an epidemic, when depletion of susceptibles is negligible, we expect the forward generation-interval distribution to be similar to the intrinsic generation-interval distribution.
As an epidemic progresses, an infector is less likely to infect individuals later in time due to decrease in susceptibles and the distribution of forward generation intervals will be shorter than the intrinsic distribution.
Conversely, when an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual, and the distribution of backward generation intervals will be shorter than the intrinsic distribution.
When epidemic is subsiding, most infections are caused by the remaining infectors, rather than new infectors, and the backward generation interval will be long \citep{champredon2015intrinsic}.

In practice, generation intervals are often measured by aggregating available information from contact tracing; while an epidemic is ongoing, measurements are ``censored'': we don't know what happens after the observation time.
These censored observations can be thought of as a weighted average of backward generation-interval distributions; like them, the observed censored intervals will be shorter than intrinsic distributions early in the epidemic.

Observed generation intervals also vary across space.
In a structured population (that is, any population that does not mix homogeneously), susceptibility will tend to decrease more quickly in the neighborhood of infected individuals than in the general population even in the early phase of an epidemic. 
This means that contacts made late in infection are more likely to be ``masked'' due to infectious contacts that were made earlier.
Therefore, realized generation intervals will be shorter and observed generation intervals may underestimate the intrinsic generation-interval distribution even when population-level susceptibility is very high.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate $r$, the reproductive number $\RR$ on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

In this study, we explore spatiotemporal variation in observed generation-interval distributions.
We show that directly using the \emph{observed} generation-interval distributions will generally underestimate the reproductive number \RR, while directly using \emph{intrinsic} generation-interval distributions (i.e., estimated from infectiousness profiles) will tend to overestimate \RR.
We discuss statistical approaches to estimating the effective generation-interval distributions and reproductive number of an outbreak from observed generation-interval distributions.
\swp{Come back and fix this}

@@

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
An individual-level intrinsic kernel $K_a(\tau)$ describes the rate at which secondary infections are expected to be caused by an infected individual $a$.
The population-level kernel is given by integrating over individual variations (e.g., variation in latent and infectious periods):
\begin{equation}
K(\tau) = \int K_a (\tau) dA.
\end{equation}
The population-level kernel describes the rate at which secondary infections are expected to be caused by an \emph{average} infected individual.
\jd{This is assuming that we know how to average across infected individuals. Do we need to assume that individual kernel properties are independent of risk of infection? If we have high-degree individuals, does it already break that assumption?}

There are two components to infection kernels: intrinsic infectiousness of an infected individual (the basic reproductive number) and time distribution of new infections (generation-interval distribution).
The basic reproductive number -- expected number of secondary cases caused by an \emph{average} primary case in a fully susceptible population -- is defined as: 
\begin{equation}
\RR_0 = \int K(t).
\end{equation}
\jd{OK, this one definitely looks problematic; that's not what happens with high degree.}
The intrinsic generation-interval distribution is simply the population-level kernel normalized by the basic reproductive number:
\begin{equation}
g(\tau) = \frac{K(t)}{\RR_0}.
\end{equation}

We assume that the disease incidence $i(\tau)$ is the product of the current infectiousness of previously infected individuals and $S(\tau)$, the proportion of the population susceptible.
\jd{Do we need to be clear when we do and don't assume this?}
\begin{equation}
i(\tau) = S(\tau) \int K(s) i(\tau-s) ds = \RR \int g(s) i(\tau-s) ds.
\end{equation}
This model, also referred to as the renewal equation, can represent a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model}.
Over a period of time where $S$ remains roughly constant, we would expect approximately exponential growth in $i$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
\begin{equation}
\frac{1}{\RR} = \int g(\tau) \exp(-r \tau) d\tau.
\end{equation}

@@

Generation intervals estimated during an outbreak are ``right-censored'', because intervals that end after the observation time are not observed. 
Several studies have associated the backward generation-interval distribution with the early-epidemic censored generation-interval distribution \citep{tomba2010some, nishiura2010time, champredon2015intrinsic, britton2019estimation}. 
The censored generation-interval distribution in fact averages over backward generation-interval distributions up until the observation time.
While we mathematically expect these underlying distributions to be the same (and thus the same distribution as their average) during the exponential growth phase, these assumptions can break down (due to susceptible depletion or other temporal change).
This, it is useful to conceptually distinguish the censored distribution the backward distribution.

The density of infections occuring at time $s$ caused by infectors who were infected at time $s-\tau$ time ago is given by
\begin{equation}
i_{s-\tau}(s) = \RR i(s-\tau) g(\tau) S(s)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, describes a distribution of infection that occured $\tau$ time units before the referece time $t$ and is proportional to $i_{t-\tau}(t)$:
\jd{Is it nicer to keep $s$ in the numerator? Otherwise, why do we bother with $s$ at all?}
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^t i(t-x) g(x) d\tau}
\end{equation}

\jd{Can we just skip this \P, or put in an appendix or something?}
On the other hand, the censored generation-interval distribution, $c_t(\tau)$, describes a distribution of \emph{all} infections that are $\tau$ time units apart from any reference time $s$ before $t$ and is proportional to $\int_0^t i_{s-\tau}(s) ds$.
Then, the censored generation-interval distribution is given by
\begin{equation}\label{eq:obsg}
c_t(\tau) = \frac{g(\tau) \int_0^t i(s-\tau) S(s) ds}{\int_0^t g(\tau) \int_0^t i(s-\tau) S(s) ds d\tau}
\end{equation}
Thus, the censored generation-interval distribution can be interpreted as a weighted distribution of the backward generation intervals:
\begin{equation}
c_t(\tau) \propto \int_0^t b_s(\tau) S(s) ds.
\end{equation}
Therefore, we expect the backward generation-interval distribution to be equivalent to the censored generation-interval distribution when $S \approx 1$.
